Reparametrization of COM-Poisson Regression Models with Applications in the Analysis of Experimental Data

### 2017-11-13 article.Rnw
Improve abstract and introduction

%-------------------------------------------
% Abstract
\Abstract{
  In the analysis of count data seldom the equidispersion assumption is
  suitable, hence the Poisson regression model is inappropriate in many
  situations. The COM-Poisson is a member of the exponential family, has
  the Poisson and geometric distributions as special cases and the
  Bernoulli distribution as a limiting case. It can deal with both under
  and overdispersed count data. Therefore it is a suitable distribution
  to the analysis of dispersed count data. In spite of the nice
  properties of COM-Poisson, the location parameter of the distribution
  does not represent the conditional mean making difficult
  interpretation in a regression structure. In this paper we present a
  very simple reparametrization based on mean approximation. The
  estimation and inference is based on likelihood and simulation study
  showed that the maximum likelihood estimators are unbiased and
  consistent for both regression and precision parameters. Furthermore,
  we observed empirical correlation between regression and precision
  parameters close to zero suggesting the orthogonality feature in the
  proposed mean reparametrized model. We present applications of the
  proposed model through three case studies with over, under and
  equidispersed counts. Results show improvements over the conventional
  approaches. Therefore, its use in the analysis of count data is
  encouraged. The computational routines for fitting of original and
  reparametrized COM-Poisson models are available in the supplementary
  material.
}

%-------------------------------------------
% Key-words
\Keywords{
COM-Poisson; Count data; Over/Under-dispersion; Likelihood inference
}

@@

Count data are variables that assume non-negative integer values and
represent the number of times an event occurs in the observation
unit. This kind of data is common in experimental research such as
the number of grains produced by a plant, number of fruits produced by a
tree, number of insects captured by a trap, among others. Since the
seminal paper by \citet{Nelder1972} introduced the class of generalized
linear models (GLM), analysis of count data often consider the Poisson
regression model. This model provides a suitable distribution for a
counting random variable and an efficient Newton scoring algorithm is
used to fit it. Furthermore, the theoretical background for GLMs is
well established in the class of dispersion models \citep{Jorgensen1997}
as a generalization of the exponential family of distributions.

In spite of the advantages of the GLM framework, the Poisson
distribution has only one parameter, which represents the mean, and
also, the variance of the count random variable. This relationship is
referred as equidispersion.  However, this assumption can be unsuitable
to many situations and when the Poisson model is applied in these cases
the standard errors for the parameters estimates are inconsistent
\citep{Winkelmann1995}.

In practice, overdispersion, when the variance greater than the mean, is
largely reported in the literature and may occur due to the absence of
relevant covariates, heterogeneity of sampling units, different domain
size not considered and excess of zero \citep{Hinde1998}. Lesser
reported, but that has received attention from the statistical
community, is a underdispersion case, when the variance is less than the
mean. The processes that reduce the variability are not as well known as
those lead to extra variability. For the same reason, there are a few
approaches to deal underdispersed cases. The explanatory mechanisms to
underdispersion may be related to Poisson process. When the time between
events is not exponential distributed, the number of events can be over
or underdispersed, this motivated the class of duration dependence
models \citep{Winkelmann1995}. Other possibility to explain
underdispersion are order statistics. Suppose $Y_i$ is Poisson
distributed then maximum of $Y_i$ is underdispersed.

The strategies for constructing alternative count distributions are
related with the causes of the non-equidispersion. Particular to
overdispersion case Poisson mixed models are widely applied. One popular
example of this approach is the negative-binomial model, where the mean
is added a random noise that is assumed to follow a gamma
distribution. However, other distributions for the random effects may be
considered, for example the model Poisson-Tweedie \citep{Bonat2017} and
Poisson-Inverse-Gaussian. For underdispersion, the Gamma-Count
distribution assumes a gamma distribution for the time between events,
thus it can deal with dispersed counts \citep{Zeviani2014}. The
COM-Poisson distribution is given by a generalization the Poisson
distribution allowed for a non-linear decrease in ratios of successive
probabilities \citep{Shmueli2005}.

The COM-Poisson is a member of the exponential family, has the Poisson
and geometric distributions as special cases and the Bernoulli
distribution as a limiting case. It can deal with both under and
overdispersed count data. Some recently applications of the COM-Poisson
distribution are related by \citet{Lord2010} for analysis of traffic
crash data, \citet{Sellers2010} to model airfreight breakage and book
purchases and \citet{Huang2017} with analysis of attendance data,
takeover birds and cotton bolls count. The main disadvantage of
COM-Poisson distribution is that the location parameter does not
represent the regression conditional mean making difficult
interpretation in a regression structure. \citet{Huang2017} proposed a
mean-parametrization COM-Poisson distribution to avoid this issue, in
this approach the mean parameter is obtained by solving the root of an
equation defined as a infinite sum.

The main goal of this article is to proposed a novel COM-Poisson
parametrization based on mean approximation given by
\citet{Shmueli2005}. In this parametrization the probability mass
function is written with $\mu$ and $\phi$ as function of original
parameters, where $\mu$ is the mean and $\phi$ is the precision or
dispersion parameter. Opposite the original parametrization, the
proposed parametrization leads to interpretability of the parameters in
a regression model and it is more simple than mean-parametrization gives
by \cite{Huang2017} because in this case it is not necessary find $\mu$
as the root of an equation involving an infinite sum. We conduce a
simulation study to performe the properties of the maximum likelihood
estimators for the proposed mean-parametrized COM-Poisson regression
model.

This paper is organized as follows. In \autoref{background} we present
the COM-Poisson distribution and \cite{Huang2017} parametrization. The
proposed reparametrization and assessement of moments approximations is
considered in the \autoref{reparametrization}. In the
\autoref{estimation-and-inference} we show the estimation and inference
methods based on likelihood function whose properties are assess in
\autoref{simulation-study} by simulations. Applications of
reparametrized COM-Poisson regression model are illustrated in
\autoref{case-studies} with three case studies. The analysis are
performed using the software R \citep{Rcore2017}. The source code and
data sets are available in supplementary material\footnote{Available on
  \texttt{\url{http://www.leg.ufpr.br/~eduardojr/papercompanions}}
  \label{papercompanion}.}.

@@

print(xy1, split = c(1, 1, 3, 1), more = TRUE)
print(xy2, split = c(2, 1, 3, 1), more = TRUE)
print(xy3, split = c(3, 1, 3, 1), more = TRUE)

>>>>

% -------------------------------------------
% Abstract
\Abstract{ In the analysis of count data often the equidispersion
  assumption is not suitable, hence the Poisson regression model is
  inappropriate.  As a generalization of the Poisson distribution the
  COM-Poisson distribution can deal with under-, equi- and overdispersed
  count data.  It is a member of the exponential family of distributions
  and has the Poisson and geometric distributions as special cases as
  well as the Bernoulli distribution as a limiting case.  In spite of
  the nice properties of the COM-Poisson distribution, its location
  parameter does not correspond to the expectation, which in turn
  becomes challenging the interpretation of the regression models
  specified using such a distribution. In this paper, we propose a
  straightforward reparametrization of the COM-Poisson distribution
  based on an approximation to the expectation of this distribution. The
  main advantage of our new parametrization is the straightforward
  interpretation of the regression coefficients in terms of the
  expectation of the count response variable, as usual in the context of
  generalized linear models. Furthermore, the estimation and inference
  for the new COM-Poisson regression model can be done based on the
  likelihood paradigm. We carried out simulation studies to verify the
  finite sample properties of the maximum likelihood estimators. The
  results from our simulation study show that the maximum likelihood
  estimators are unbiased and consistent for both regression and
  dispersion parameters. We observed that the empirical correlation
  between the regression and dispersion parameter estimators is close to
  zero, which suggests that these parameters are orthogonal. We
  illustrate the application of the proposed model through the analysis
  of three data sets with over-, under- and equidispersed count data.
  The study of distribution properties through a consideration of
  dispersion, zero-inflated and heavy tail indices, and the results of
  data analysis show the flexibility over standard
  approaches. Therefore, we encourage the application of the new
  parametrization for the analysis of count data in the context of
  COM-Poisson regression models.  The computational routines for fitting
  the original and new version of the COM-Poisson regression model and
  the analyzed data sets are available in the supplementary material.
}

%-------------------------------------------
% Key-words
\Keywords{ COM-Poisson, Overdispersion, Underdispersion, Count data,
  Likelihood inference.  }

@@

Count data are random variables that assume non-negative integer values
and represent the number of times an event occurs in the observation
unit. This kind of data is common in crop sciences such as the number of
grains produced by a plant, number of fruits produced by a tree, number
of insects captured by a trap, to cite but a few.  Since the seminal
paper of \citet{Nelder1972} where the class of the generalized linear
models (GLMs) was introduced, the analysis of count data often employs
the Poisson regression model. This model provides a suitable strategy
for the analysis of count data and an efficient Newton scoring algorithm
can be used for fitting the model.

In spite of the advantages of the Poisson regression model, the Poisson
distribution has only one parameter, which represents both the
expectation and variance of the count random variable.  This restriction
in the relationship between the expectation and variance induced by the
Poisson distribution is referred as equidispersion.  However, in
practical data analysis such a restriction can be unsuitable, since the
observed data can present both the variance smaller or larger than the
expectation, leading to the cases of under and overdispersion,
respectively.  The main problem of the application of the Poisson
regression model to non equidispersed count data is that the standard
errors associated with the regression coefficients are inconsistently
estimated, which in turn can lead to misleading
inferences~\citep{Winkelmann1995, Bonat2017}.

In practice, overdispersion is largely reported in the literature and
may occur due to the absence of relevant covariates, heterogeneity of
sampling units, different domain size not considered in the analysis and
excess of zeros~\citep{Hinde1998}.  The case of underdispersion is less
report in the literature, however, it has been of increasing interest in
the statistical community. The processes that reduce the variability are
not as well-known as those leading to extra variability.  For the same
reason, there are a few approaches to deal with underdispersed count
data. The explanatory mechanisms leading to underdispersion may be
related to the underline Poisson process generating the count data.
% Here, you gave the idea that you are going to explain all mechanisms.
% Perhaps, you can improve on it!! Have a look at my material for
% regression models school.
When the time between events is not exponential distributed, the number
of events can be over or underdispersed, this process motivated the
class of duration dependence models~\citep{Winkelmann1995}.  Another
possibility to explain the underdispersion phenomenon is order
statistics.  Suppose $Y_i$ is Poisson distributed then the maximum of
$Y_i$ is underdispersed. (NEED REF).

The strategies for constructing alternative count distributions are
related with the causes of the non-equidispersion. In special to the
overdispersion case the Poisson mixture models are widely applied.  One
popular example of this approach is the negative-binomial model, where a
gamma random effect is added to the expectation of the Poisson
distribution.  However, other distributions can be used to represent the
random effects.  For example the Poisson-Tweedie model~\citep{Bonat2017}
and its special cases as the Poisson inverse-Gaussian and Neyman-Type A
consider that the random effects are Tweedie, inverse Gaussian and
Poisson distributed, respectively.  The Gamma-Count distribution assumes
a gamma distribution for the time between events, thus it can deal with
underdispersed as well as overdispersed count
data~\citep{Zeviani2014}. The COM-Poisson distribution is obtained by a
generalization of the Poisson distribution allowing for a non-linear
decrease in the ratios of successive probabilities~\citep{Shmueli2005}.

The COM-Poisson distribution is a member of the exponential family and
it has the Poisson and geometric distributions as special cases as well
as the Bernoulli distribution as a limiting case. It can deal with both
under and overdispersed count data.  Some recently applications of the
COM-Poisson distribution include \citet{Lord2010} for the analysis of
traffic crash data, \citet{Sellers2010} for the modelling of airfreight
breakage and book purchases and \citet{Huang2017} to the analysis of
attendance data, takeover birds and cotton bolls count.  The main
disadvantage of the COM-Poisson regression model as presented
in~\citet{Sellers2010} is that its location parameter does not
correspond to the expectation of the distribution, which in turn becomes
the interpretation of the regression coefficients challenging and no
comparable with standard approaches as the Poisson and negative binomial
regression models. \citet{Huang2017} proposed a mean-parametrization of
the COM-Poisson distribution in order to avoid such an issue.  In this
approach the mean parameter is obtained by solving an non-linear
equation defined as an infinite sum.  Consequently, it is
computationally demanding and liable to numerical problems.

The main goal of this article is to propose a novel COM-Poisson
parametrization based on the mean approximation presented by
\citet{Shmueli2005}. In this parametrization, the probability mass
function is written in terms of $\mu$ and $\phi$, where $\mu$ is the
expectation and $\phi$ is the dispersion parameter. In contrast to the
original parametrization, the proposed parametrization leads to
regression coefficients directly associated with the expectation of the
response variable as usual in the context of generalized linear models.
Consequently, the obtained regression coefficients are comparable with
the ones obtained by standard approaches as the Poisson and negative
binomial regression models.  Furthermore, our novel COM-Poisson
parametrization is simpler than the strategy proposed
by~\cite{Huang2017}, since it does not involve any numerical method for
solving non-linear equations.

This paper is organized as follows. In \autoref{background} we present
the COM-Poisson distribution and the strategy proposed
by~\cite{Huang2017}.  The proposed reparametrization, assessment of
moment approximations, and study of distribution properties are
considered in the~\autoref{reparametrization}. In
the~\autoref{estimation-and-inference} we present the estimation and
inference for the novel COM-Poisson regression model based on the
likelihood paradigm. The properties of the maximum likelihood estimators
are assess in~\autoref{simulation-study} through simulation studies.  We
illustrate the application of the new COM-Poisson regression model
through the analysis of three data sets. We provide an \texttt{R}
implementation of the COM-Poisson and reparameterized COM-Poisson
regression models as well as the analyzed data sets in the supplementary
material.\footnote{Available on
  \texttt{\url{http://www.leg.ufpr.br/~eduardojr/papercompanions}}
  \label{papercompanion}.}.

@@

print(xy1, split = c(1, 1, 3, 1), more = TRUE)
print(xy2, split = c(2, 1, 3, 1), more = TRUE)
print(xy3, split = c(3, 1, 3, 1), more = FALSE)
